# Visualize the dendrogram with the help of ggplot and ggtree

- https://4va.github.io/biodatasci/r-ggtree.html#tree_import
- geom_point2()
  - https://rdrr.io/bioc/ggtree/man/geom_point2.html
- Phylogenetic Tree Annotation
  - https://yulab-smu.top/treedata-book/chapter5.html
- Mapping Data to The tree Structure
  - https://yulab-smu.top/treedata-book/chapter7.html
## Libraries

```{r}
library(dplyr)
library(tidyr)
library(stringr)
library(ggplot2)
library(ggrepel)
if (!require("BiocManager", quietly = TRUE)){
    install.packages("BiocManager")
    BiocManager::install("ggtree")  
}
# for plotting the dendrogram
library(ggtree)

# for interaction
library(plotly)
```


## Constants

```{r}
# Par
# Path
# - Please change the root directory of the project to your preference
PATH_ROOT <- "~/CS765_final_project"
PATH_ATUSSUM_RAW_DATA <- file.path(PATH_ROOT, "data")
PATH_TREE <- file.path(PATH_ROOT, "data")
PATH_PLOT <- file.path(PATH_ROOT, "presentation")
dir.create(PATH_ATUSSUM_RAW_DATA)
dir.create(PATH_TREE)
dir.create(PATH_TREE)
dir.create(PATH_PLOT)

# File
FILE_ATUSSUM_RAW_DATA <- file.path(PATH_ATUSSUM_RAW_DATA, "atussum_0321.csv")
FILE_TREE_A <- file.path(PATH_TREE, "tree_newick_a.nwk")
FILE_TREE_B <- file.path(PATH_TREE, "tree_newick_b.nwk")
FILE_PLOT_A_WITH_TEXT <- file.path(PATH_PLOT, "plot_a_with_text.png")
FILE_PLOT_A_NO_TEXT <- file.path(PATH_PLOT, "plot_a_no_text.png")
FILE_PLOT_A_HTML <- file.path(PATH_PLOT, "plotly_dendrogram_interactive_a.html")
FILE_PLOT_B_WITH_TEXT <- file.path(PATH_PLOT, "plot_b_with_text.png")
FILE_PLOT_B_NO_TEXT <- file.path(PATH_PLOT, "plot_b_no_text.png")
FILE_PLOT_B_HTML <- file.path(PATH_PLOT, "plotly_dendrogram_interactive_b.html")
```

# Real Data

## Read in the data

```{r}
df_real <- read.csv(FILE_ATUSSUM_RAW_DATA)
```

### Preprocess

```{r}
# filter out the activities
df_real <-
  df_real %>%
  select(!matches(match = "t\\d+"))

# retain only the variables of interest
df_real <-
  df_real %>%
  select(
    id = TUCASEID,
    employ_status = TELFS,
    sex = TESEX,
    age = TEAGE,
    num_children = TRCHILDNUM
    # income = HUFAMINC
  )

# bin the continuous variables
df_real <-
  df_real %>%
  mutate(
    new_employ_status = factor(case_when(
      employ_status %in% c(1, 2) ~ "E1",
      employ_status %in% c(3, 4) ~ "E2",
      employ_status %in% c(5) ~ "E3"
    )),
    new_age = factor(case_when(
      age < 30 ~ "A1",
      age >= 30 & age <= 65 ~ "A2",
      age > 65 ~ "A3"
    )),
    new_num_children = factor(case_when(
      num_children == 0 ~ "C1",
      num_children > 0 ~ "C2",
      # num_children >= 3 ~ "C3"
    )),
    new_sex = factor(case_when(
      sex == 1 ~ "S1",
      sex == 2 ~ "S2"
    ))
  )
```

## design-a) sex-age-employ

### summarize the group distributions

```{r}
# iterative histogram (the proportion of people at each level)
list_group_vars_a <- c("new_sex", "new_age", "new_employ_status")
# list_group_vars_a <- c("new_age", "new_sex", "new_employ_status")

list_df_hist_a <- list()

for (i_group_vars in 1:length(list_group_vars_a)) {
  group_vars <- list_group_vars_a[1:i_group_vars]

  # the level in the tree (root is 1)
  i_level <- i_group_vars + 1
  df_hist_this_level <-
    df_real %>%
    group_by(across(all_of(group_vars))) %>%
    summarise(
      count = n(),
      prop = 100 * count / nrow(df_real)
    ) %>%
    mutate(i_level = i_level)

  df_hist_this_level$node_label <-
    apply(df_hist_this_level[, group_vars], 1, function(row) {
      paste(row, collapse = "")
    })
  list_df_hist_a[[i_group_vars]] <- df_hist_this_level
}
df_hist_a <- bind_rows(list_df_hist_a)

# add the root node
df_hist_a <-
  df_hist_a %>%
  add_row(node_label = "All", prop = 100, count = nrow(df_real))
```

### Generate the tree features

```{r}
COLOR_ALL <- "#8c8c8c"
COLOR_MALE <- rgb(0, 0, 255, 128, maxColorValue = 255)
COLOR_FEMALE <- rgb(255, 64, 64, 128, maxColorValue = 255)
COLOR_YOUNG <- rgb(255, 215, 0, 128, maxColorValue = 255)
COLOR_MIDDLE <- rgb(0, 201, 87, 128, maxColorValue = 255)
COLOR_OLD <- rgb(153, 50, 204, 128, maxColorValue = 255)
COLOR_NO_CHILD <- rgb(139, 62, 47, 128, maxColorValue = 255)
COLOR_WITH_CHILD <- rgb(127, 255, 212, 128, maxColorValue = 255)
COLOR_EMP <- rgb(255, 128, 0, 128, maxColorValue = 255)
COLOR_UMP <- rgb(0, 245, 255, 128, maxColorValue = 255)
COLOR_NLF <- rgb(255, 0, 255, 128, maxColorValue = 255)

# COLOR_ALL <- "#404040"
# COLOR_MALE <- "#8faadc"
# COLOR_FEMALE <- "#f4b183"
# COLOR_MALE_YOUNG <- "#b4c7e7"
# COLOR_MALE_MIDDLE <- COLOR_MALE
# COLOR_MALE_OLD <- "#2f5597"
# COLOR_FEMALE_YOUNG <- "#f8cbad"
# COLOR_FEMALE_MIDDLE <- COLOR_FEMALE
# COLOR_FEMALE_OLD <- "#c55a11"

# RGB(0,0,255)', 'RGB(255,64,64)', 'RGB(255,215,0)', 'RGB(0,201,87)', 'RGB(153,50,204)', 'RGB(255,128,0)', 'RGB(0,245,255)', 'RGB(255,0,255)', 'RGB(139,62,47)', 'RGB(127,255,212)', 'RGB(255,235,205)', 'RGB(151,255,255)', 'RGB(255,182,193)', 'RGB(25,25,112)', 'RGB(105,139,34)', 'RGB(139,0,0)', 'RGB(139,139,0)', 'RGB(56,142,142)', 'RGB(128,128,105)', 'RGB(120,120,120)
SIZE_SCALING_FACTOR <- 1
df_node_feature_a <-
  df_hist_a %>%
  mutate(
    node_size = prop * SIZE_SCALING_FACTOR,
    node_label_abrv = str_extract(
      string = node_label,
      pattern = "(All)|\\w\\d+$"
    ),
    node_prop_abrv = round(prop, 1),
    node_label_readable = case_when(
      node_label_abrv == "All" ~ "All",
      node_label_abrv == "S1" ~ "Male",
      node_label_abrv == "S2" ~ "Female",
      node_label_abrv == "A1" ~ "<30",
      node_label_abrv == "A2" ~ "[30,65]",
      node_label_abrv == "A3" ~ ">65",
      node_label_abrv == "E1" ~ "E",
      node_label_abrv == "E2" ~ "U",
      node_label_abrv == "E3" ~ "N"
      # node_label_abrv == "E1" ~ "EMP",
      # node_label_abrv == "E2" ~ "UMP",
      # node_label_abrv == "E3" ~ "NLF",
    ),
    node_label_full = case_when(
      node_label_abrv == "All" ~ "All",
      node_label_abrv == "S1" ~ "Male",
      node_label_abrv == "S2" ~ "Female",
      node_label_abrv == "A1" ~ "<30 years",
      node_label_abrv == "A2" ~ "[30,65] years",
      node_label_abrv == "A3" ~ ">65 years",
      node_label_abrv == "E1" ~ "Employed",
      node_label_abrv == "E2" ~ "Unemployed",
      node_label_abrv == "E3" ~ "Not in Labor Force"
    ),
    node_text = paste0(
      node_label_readable,
      "\n",
      node_prop_abrv,
      " %"
    ),
    node_color = case_when(
      node_label == "All" ~ COLOR_ALL,
      grepl(pattern = "S1$", x = node_label) ~ COLOR_MALE,
      grepl(pattern = "S2$", x = node_label) ~ COLOR_FEMALE,
      grepl(pattern = "A1$", x = node_label) ~ COLOR_YOUNG,
      grepl(pattern = "A2$", x = node_label) ~ COLOR_MIDDLE,
      grepl(pattern = "A3$", x = node_label) ~ COLOR_OLD,
      grepl(pattern = "E1$", x = node_label) ~ COLOR_EMP,
      grepl(pattern = "E2$", x = node_label) ~ COLOR_UMP,
      grepl(pattern = "E3$", x = node_label) ~ COLOR_NLF
    )
  ) %>%
  select(
    node_label,
    node_label_abrv,
    node_label_readable,
    node_label_full,
    node_prop_abrv,
    node_size,
    node_text,
    node_color
  )
```

### Generate the tree structure

```{r}
# create the string programmatically
list_name_levels <- list(
  c("All"),
  # sex
  c("S1", "S2"),
  # age
  c("A1", "A2", "A3"),
  # employment
  c("E1", "E2", "E3")
  # c("C1","C2","C3")
)
str_tree <- create_tree_string_newick(list_name_levels)
write.table(str_tree, FILE_TREE_A, col.names = F, row.names = F, quote = F)
```

### Read files

```{r}
tree_a <- read.tree(FILE_TREE_A)
```

### Plot: Area+Length encoding (Rectangle)

```{r}
p_a <-
  # ggtree(tree_a, branch.length = "none", layout = "ellipse") %<+%
  ggtree(tree_a, branch.length = "none", layout = "dendrogram") %<+%
  # ggtree(tree_a, branch.length = "none", layout = "circular") %<+%
  # ggtree(tree_a, branch.length = "none", layout = "fan",open.angle = 120) %<+%
  df_node_feature_a +
  geom_tile(
    aes(
      height = node_size * 0.05,
      fill = node_color
    ),
    width = 0.15
    # colour = "grey50",
    # alpha = 0.5
  ) +
  scale_fill_identity() +
  # geom_point(aes(size = node_size), shape = "square",alpha=0.5) +
  # geom_text(aes(label = node_text), hjust = -0.2, vjust = 1) +
  scale_x_continuous(expand = c(0.1, 0.1)) +
  scale_y_continuous(expand = c(0.1, 0.1)) +
  # geom_text_repel(aes(label = node_text), hjust = -0.2) +
  # ensure the radius encodes the size
  scale_size_identity(
    # guide = "legend"
    # breaks = sqrt(c(5,10,20,40)),
    # labels = c(5,10,20,40)
  ) +
  layout_dendrogram() +
  # coord_flip(clip = 'off')+
  # coord_cartesian(clip = 'off') +
  theme(
    legend.position = "none"
  )
```

```{r}
g <- p_a +
  geom_text(aes(label = node_text), hjust = -0.2, vjust = 1)
ggsave(plot = g, filename = FILE_PLOT_A_WITH_TEXT, width = 30.81, height = 19.03, units = "cm")
g
```

```{r}
ggsave(
  plot = p_a,
  filename = FILE_PLOT_A_NO_TEXT,
  width = 30.81, height = 19.03, units = "cm"
)
p_a
```

#### Plotly

```{r}
plotly_a <-
  ggplotly(
    p_a +
      annotate("text",
        x = -2,
        y = -3, size = ANNOTATE_TEXT_SIZE,
        label = "Sex"
      ) +
      annotate("text",
        x = -1,
        y = -3, size = ANNOTATE_TEXT_SIZE,
        label = "Age"
      ) +
      annotate("text",
        x = 0,
        y = -3, size = ANNOTATE_TEXT_SIZE,
        label = "Employment\nStatus"
      ) +
      geom_text(
        aes(
          label = paste0(
            node_label_readable, "\n",
            node_prop_abrv
          ),
          text = paste(
            node_label_full, "\n",
            node_prop_abrv, "%"
          )
        ),
        hjust = -0.2, vjust = 1
      ),
    tooltip = c("text")
  )
plotly_a
```

#### Upload the plot to chart-studtio

- https://plotly.com/r/getting-started-with-chart-studio/

```{r}
Sys.setenv("plotly_username" = "vimchiz")
Sys.setenv("plotly_api_key" = "k3x5GH7VVwJ6LClgE7FX")
# Suppress auto open behavior:
options(browser = "false")
```

- https://chart-studio.plotly.com/~vimchiz/42.embed

```{r}
api_create(plotly_a, filename = "CS765_plotly_a")
```

#### Save as a local HTML file

```{r}
htmlwidgets::saveWidget(as_widget(plotly_a), FILE_PLOT_A_HTML)
```

## design-b) sex-age-children-employ

### summarize the group distributions

```{r}
# iterative histogram (the proportion of people at each level)
list_group_vars_b <- c("new_sex", "new_age", "new_num_children", "new_employ_status")

list_df_hist_b <- list()

for (i_group_vars in 1:length(list_group_vars_b)) {
  group_vars <- list_group_vars_b[1:i_group_vars]

  # the level in the tree (root is 1)
  i_level <- i_group_vars + 1
  df_hist_this_level <-
    df_real %>%
    group_by(across(all_of(group_vars))) %>%
    summarise(
      count = n(),
      prop = 100 * count / nrow(df_real)
    ) %>%
    mutate(i_level = i_level)

  df_hist_this_level$node_label <-
    apply(df_hist_this_level[, group_vars], 1, function(row) {
      paste(row, collapse = "")
    })
  list_df_hist_b[[i_group_vars]] <- df_hist_this_level
}
df_hist_b <- bind_rows(list_df_hist_b)

# add the root node
df_hist_b <-
  df_hist_b %>%
  add_row(node_label = "All", prop = 100, count = nrow(df_real))
```

### Generate the tree features

```{r}
COLOR_ALL <- "#8c8c8c"
COLOR_MALE <- rgb(0, 0, 255, 128, maxColorValue = 255)
COLOR_FEMALE <- rgb(255, 64, 64, 128, maxColorValue = 255)
COLOR_YOUNG <- rgb(255, 215, 0, 128, maxColorValue = 255)
COLOR_MIDDLE <- rgb(0, 201, 87, 128, maxColorValue = 255)
COLOR_OLD <- rgb(153, 50, 204, 128, maxColorValue = 255)
COLOR_NO_CHILD <- rgb(139, 62, 47, 128, maxColorValue = 255)
COLOR_WITH_CHILD <- rgb(127, 255, 212, 128, maxColorValue = 255)
COLOR_EMP <- rgb(255, 128, 0, 128, maxColorValue = 255)
COLOR_UMP <- rgb(0, 245, 255, 128, maxColorValue = 255)
COLOR_NLF <- rgb(255, 0, 255, 128, maxColorValue = 255)

COLOR_MALE_YOUNG <- COLOR_YOUNG
COLOR_MALE_MIDDLE <- COLOR_MIDDLE
COLOR_MALE_OLD <- COLOR_OLD
COLOR_FEMALE_YOUNG <- COLOR_YOUNG
COLOR_FEMALE_MIDDLE <- COLOR_MIDDLE
COLOR_FEMALE_OLD <- COLOR_OLD

SIZE_SCALING_FACTOR <- 1

# DF_LABEL_DECODE = data.frame(
#   raw_code = c("All",
#                "S1","S2",
#                "A1","A2","A3",
#                "C1","C2",
#                "E1","E2","E3"),
#   decoded_code = c("All",
#                    "Male","Female",
#                    "<30","[30,65]",">65",
#                    "0",">0",
#                    "E","U","N")
# )
df_node_feature_b <-
  df_hist_b %>%
  mutate(
    node_size = prop * SIZE_SCALING_FACTOR,
    node_label_abrv = str_extract(
      string = node_label,
      pattern = "(All)|\\w\\d+$"
    ),
    node_prop_abrv = round(prop, 1),
    node_label_readable = case_when(
      node_label_abrv == "All" ~ "All",
      node_label_abrv == "S1" ~ "Male",
      node_label_abrv == "S2" ~ "Female",
      node_label_abrv == "A1" ~ "<30",
      node_label_abrv == "A2" ~ "[30,65]",
      node_label_abrv == "A3" ~ ">65",
      node_label_abrv == "C1" ~ "0",
      node_label_abrv == "C2" ~ ">0",
      node_label_abrv == "E1" ~ "E",
      node_label_abrv == "E2" ~ "U",
      node_label_abrv == "E3" ~ "N"
      # node_label_abrv == "E1" ~ "EMP",
      # node_label_abrv == "E2" ~ "UMP",
      # node_label_abrv == "E3" ~ "NLF",
    ),
    node_label_full = case_when(
      node_label_abrv == "All" ~ "All",
      node_label_abrv == "S1" ~ "Male",
      node_label_abrv == "S2" ~ "Female",
      node_label_abrv == "A1" ~ "<30 years",
      node_label_abrv == "A2" ~ "[30,65] years",
      node_label_abrv == "A3" ~ ">65 years",
      node_label_abrv == "C1" ~ "0 child",
      node_label_abrv == "C2" ~ ">0 children",
      node_label_abrv == "E1" ~ "Employed",
      node_label_abrv == "E2" ~ "Unemployed",
      node_label_abrv == "E3" ~ "Not in Labor Force"
    ),
    node_text = paste0(
      node_label_readable,
      "\n",
      node_prop_abrv,
      " %"
    ),
    node_color = case_when(
      node_label == "All" ~ COLOR_ALL,
      node_label == "S1" ~ COLOR_MALE,
      node_label == "S2" ~ COLOR_FEMALE,
      grepl(pattern = "A1$", x = node_label) ~ COLOR_YOUNG,
      grepl(pattern = "A2$", x = node_label) ~ COLOR_MIDDLE,
      grepl(pattern = "A3$", x = node_label) ~ COLOR_OLD,
      grepl(pattern = "C1$", x = node_label) ~ COLOR_NO_CHILD,
      grepl(pattern = "C2$", x = node_label) ~ COLOR_WITH_CHILD,
      grepl(pattern = "E1$", x = node_label) ~ COLOR_EMP,
      grepl(pattern = "E2$", x = node_label) ~ COLOR_UMP,
      grepl(pattern = "E3$", x = node_label) ~ COLOR_NLF
    )
  ) %>%
  select(
    node_label,
    node_label_abrv,
    node_label_readable,
    node_label_full,
    node_prop_abrv,
    node_size,
    node_text,
    node_color
  )
```

### Generate the tree structure

```{r}
# create the string programmatically
list_name_levels <- list(
  c("All"),
  # sex
  c("S1", "S2"),
  # age
  c("A1", "A2", "A3"),
  # number of children
  c("C1", "C2"),
  # employment status
  c("E1", "E2", "E3")
)
str_tree <- create_tree_string_newick(list_name_levels)
write.table(str_tree, FILE_TREE_B, col.names = F, row.names = F, quote = F)
```

### Read files

```{r}
tree_b <- read.tree(FILE_TREE_B)
```

### Plot: Area+Length encoding (Rectangle)

```{r}
p_b <-
  # ggtree(tree_a, branch.length = "none", layout = "ellipse") %<+%
  ggtree(tree_b, branch.length = "none", layout = "dendrogram") %<+%
  # ggtree(tree_a, branch.length = "none", layout = "circular") %<+%
  # ggtree(tree_a, branch.length = "none", layout = "fan",open.angle = 120) %<+%
  df_node_feature_b +
  geom_tile(
    aes(
      height = node_size * 0.10,
      fill = node_color
    ),
    width = 0.15
    # colour = "grey50",
    # alpha = 0.5
  ) +
  scale_fill_identity() +
  # geom_point(aes(size = node_size), shape = "square",alpha=0.5) +
  # geom_text(aes(label = node_text), hjust = -0.2, vjust = 1) +
  scale_x_continuous(expand = c(0.1, 0.1)) +
  scale_y_continuous(expand = c(0.1, 0.1)) +
  # geom_text_repel(aes(label = node_text), hjust = -0.2) +
  # ensure the radius encodes the size
  scale_size_identity(
    # guide = "legend"
    # breaks = sqrt(c(5,10,20,40)),
    # labels = c(5,10,20,40)
  ) +
  layout_dendrogram() +
  # coord_flip(clip = 'off')+
  # coord_cartesian(clip = 'off') +
  theme(
    legend.position = "none"
  )
```

```{r}
ANNOTATE_TEXT_SIZE <- 5
g_b_annotated <-
  p_b +
  geom_text(aes(label = node_text), hjust = -0.2, vjust = 1) +
  annotate("text",
    x = -3,
    y = -3, size = ANNOTATE_TEXT_SIZE,
    label = "Sex"
  ) +
  annotate("text",
    x = -2,
    y = -3, size = ANNOTATE_TEXT_SIZE,
    label = "Age"
  ) +
  annotate("text",
    x = -1,
    y = -3, size = ANNOTATE_TEXT_SIZE,
    label = "# Children"
  ) +
  annotate("text",
    x = 0,
    y = -3, size = ANNOTATE_TEXT_SIZE,
    label = "Employment\nStatus"
  )
ggsave(
  plot = g_b_annotated,
  filename = FILE_PLOT_B_WITH_TEXT,
  width = 30.81 * 1.5, height = 19.03 * 1.5, units = "cm"
)
g_b_annotated
```

```{r}
p_b +
  geom_text(aes(label = node_label_readable), hjust = -0.2, vjust = 1)
```

```{r}
p_b
```

#### Plotly

```{r}
# ggplotly(p_b +
#  geom_text(aes(label = node_text), hjust = -0.2, vjust = 1)
#  ,tooltip="size")
plotly_b <-
  ggplotly(
    p_b +
      annotate("text",
        x = -3,
        y = -3, size = ANNOTATE_TEXT_SIZE,
        label = "Sex"
      ) +
      annotate("text",
        x = -2,
        y = -3, size = ANNOTATE_TEXT_SIZE,
        label = "Age"
      ) +
      annotate("text",
        x = -1,
        y = -3, size = ANNOTATE_TEXT_SIZE,
        label = "# Children"
      ) +
      annotate("text",
        x = 0,
        y = -3, size = ANNOTATE_TEXT_SIZE,
        label = "Employment\nStatus"
      ) +
      geom_text(
        aes(
          label = paste0(
            node_label_readable, "\n",
            node_prop_abrv
          ),
          text = paste(
            node_label_full, "\n",
            node_prop_abrv, "%"
          )
        ),
        hjust = -0.2, vjust = 1
      ),
    tooltip = c("text")
  )
plotly_b
```

#### Upload the plot to chart-studtio

- https://plotly.com/r/getting-started-with-chart-studio/

```{r}
Sys.setenv("plotly_username" = "vimchiz")
Sys.setenv("plotly_api_key" = "k3x5GH7VVwJ6LClgE7FX")
# Suppress auto open behavior:
options(browser = "false")
```

- https://chart-studio.plotly.com/~vimchiz/40.embed

```{r}
api_create(plotly_b, filename = "CS765_plotly_b")
```

#### Save as a local HTML file

```{r}
htmlwidgets::saveWidget(as_widget(plotly_b), FILE_PLOT_B_HTML)
```
