
## Import

```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
library(scales)
```

## Read in the data

```{r}
df = read.csv("aiddata_reduced.csv")
```

Question 1: a) How does the amount donated vs. amount received change over time for each country?; b) Are there countries that mostly send or mostly receive and countries that have a similar amount of donations they receive and send?; c) Are there countries that change their role over time? That is, they used to mostly send donations and turn into mostly receiving donations and vice-versa?; d) Are there countries in which you can find a sudden increase (“peak”) or a sudden decrease (“valley”)?

```{r}
# convert to long format
df_long = 
  df%>%
    pivot_longer(cols = c("donor","recipient"),
                 names_to = "role",
                 values_to = "country")
```

```{r}
df_long%>%
  group_by(country)%>%
  summarise(
    n_donate=sum(role=="donor"),
    n_receive=sum(role=="recipient")
    )
```

```{r}
# a) How does the amount donated vs. amount received change over time for each country?
df_long %>%
  # filter(country=="Korea")%>%
  ggplot(aes(x=year,y=commitment_amount_usd_constant,color=role))+
  geom_line()+
  geom_point()+
  facet_wrap(.~country)+
  scale_y_continuous(
    trans = log10_trans(),
    breaks = trans_breaks("log10", function(x) 10^x),
    labels = trans_format("log10", math_format(10^.x))
    )
```





